---
title: "project"
format: html
editor: visual
---

## EVA(Covarites)

```{r}
diabetes <- read.csv('diabetes_dataset.csv')
```

```{r}
str(diabetes)
summary(diabetes)
```

To explore how individual covariates may be associated with the probability of being diagnosed with diabetes, we conducted exploratory visualizations and statistical summaries. The covariates analyzed include **glucose level**, **HbA1c level**, **family history of diabetes**, and **medication use**.

```{r}
diabetes <- as.data.frame(diabetes)
boxplot(Glucose ~ Outcome,
        data = diabetes,
        main = "Glucose Level by Diabetes Status",
        xlab = "Diabetes (0 = No, 1 = Yes)",
        ylab = "Glucose Level",
        col = c("lightblue", "salmon"))
```

This pattern aligns with clinical understanding, where elevated blood glucose is a primary indicator of diabetes. The visual evidence strongly supports the inclusion of glucose level as a key predictor in modeling diabetes diagnosis.

```{r}
boxplot(HbA1c ~ Outcome,
        data = diabetes,
        main = "HbA1c Level by Diabetes Status",
        xlab = "Diabetes (0 = No, 1 = Yes)",
        ylab = "HbA1c Level",
        col = c("lightblue", "salmon"))

```

Individuals diagnosed with diabetes tend to have significantly higher **glucose** and **HbA1c** levels, are more likely to have a **family history of diabetes**, and report **greater medication use**, highlighting all four as strong indicators of disease presence.

```{r}
barplot(prop.table(table(diabetes$FamilyHistory, diabetes$Outcome), 1),
        beside = TRUE,
        col = c("lightgreen", "orange"),
        legend = c("No Diabetes", "Diabetes"),
        names.arg = c("No Family History", "Family History"),
        main = "Proportion of Diabetes by Family History",
        ylab = "Proportion")

```

The barplot shows that individuals **without a family history** of diabetes are overwhelmingly non-diabetic, while those **with a family history** are predominantly diagnosed with diabetes. This strong contrast suggests that family history is a powerful categorical predictor of diabetes risk.

```{r}
barplot(prop.table(table(diabetes$MedicationUse, diabetes$Outcome), 1),
        beside = TRUE,
        col = c("lightgreen", "orange"),
        legend = c("No Diabetes", "Diabetes"),
        names.arg = c("No Medication", "Using Medication"),
        main = "Proportion of Diabetes by Medication Use",
        ylab = "Proportion")

```

The barplot shows that individuals **using medication** have a higher proportion of diabetes compared to those **not on medication**, who are more likely to be non-diabetic. This pattern reflects expected treatment behavior, where medication use is closely linked to a diabetes diagnosis.
